Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-04-17 acm/acm_paper_submitted.tex
Updated paper. Slightly larger figures, added some space below some figures. Column width check.
pdfltx acm_paper_submitted.tex

%\documentclass[10pt,conference,compsocconf]{IEEEtran}
%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}
\input{support/macros} % color is defined in macros or misc_mac
\input{support/misc_mac}
\input{support/setupicase}

@@

\section{Introduction and Motivation for RBF-FD}
\label{sec:rbf}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.60\linewidth]{figures/RBFStencil_n32_a.pdf}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=32$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 32 nonzero entries per row.}
\end{figure}

@@

\begin{figure*}[tbh]
  \centering
  \subfigure[Best case.]{\includegraphics[width=.28\linewidth]{figures/flops_to_bytes_best-crop.pdf}\label{fig:ratio_best}}%
  \subfigure[Worst case.]{\includegraphics[width=.28\linewidth]{figures/flops_to_bytes_worst-crop.pdf}\label{fig:ratio_worst}}%
  \subfigure[Worst case (No cacheline effects)]{\includegraphics[width=.28\linewidth]{figures/flops_to_bytes_no_cache-crop.pdf}\label{fig:ratio_worst_nocache}}

@@

\begin{figure}[t]
  \centering 
  \includegraphics[width=.9\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a bandwidth from memory to computational units of 150~GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
\end{figure}

@@

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first array is {\tt col\_id} that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists
of $n_z$ entries that label the column index of each nonzero in
$A$. %***
The other array is  {\tt data} that gives the values of
the nonzeros in the four matrices. The values of the matrices are
interleaved by groups of four so that the value $A^l_{col\_id[k]}$ of
the $k$th nonzero of the $l$th matrix is at index $\floor{\frac{k-1}{4}}*4*n_m + (l-1)*4+(k-1)\%4$.
(In other words, the first four entries are from $A^{1}$ and the 
next four are from $A^2$.) The vectors {\tt x} are
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly 
grouped in sets of 16, which correspond to four derivatives applied to
four functions, at a single point. %\ger{All matrices are stored as one-dimensional vectors and pointers to individual rows and elements are explicitely computed.}
\looseness=-1

@@

  \caption{Bandwidth performance under idealized conditions.  Entries
    with "cpp" (triangles) denote cases where vectorization is
    compiler generated.}
  \label{fig:bandwidth}
  \label{fig:band_rw_MIC}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=.9\linewidth]{figures/bandwidth_gather_unpack.pdf}
  \caption{Performance of {\tt gather} and {\tt unpack} operations.}
  \label{fig:band_gather} 
\end{figure}

@@

\begin{figure*}
% Figure 9
  \centering
% was width=.48
  \subfigure[Supercompact and Random.]{\includegraphics[width=.35\linewidth]{figures/random_supercompact.pdf}\label{fig:perf_ideal}}
% was width=.48
  \subfigure[Compact and RBF-FD.]{\includegraphics[width=.35\linewidth]{figures/rbf_compact.pdf}\label{fig:perf_realistic}}

@@

  \includegraphics[width=.8\linewidth]{figures/mic_performance_nb_threads.pdf}

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical "one-matrix one-vector" SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.
    We also show peak theoretical performance for v1m1 and v4m1 on the MIC. 
    Not shown is the peak theoretical performance of 210 Gflops for v4m4. 
    }
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

@@

\section*{Acknowledgment}
Erlebacher acknowledges funding from NSF grant
DMS-\#0934331 (FSU), and the use of the MIC at FSU and at LCSE/U. Minnesota (provided by Prof. D. Yuen). NCAR is sponsored by NSF. 
Flyer acknowledges support of NSF grant DMS-0934317.

>>>>

%\documentclass[10pt,conference,compsocconf]{IEEEtran}
%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}
\input{support/macros} % color is defined in macros or misc_mac
\input{support/misc_mac}
% resets margins. Bad. 
%\input{support/setupicase}

@@

\pagebreak
\section{Introduction and Motivation \\for RBF-FD}
\label{sec:rbf}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.80\linewidth]{figures/RBFStencil_n32_a.pdf}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=32$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 32 nonzero entries per row.}
\vspace{3ex}
\end{figure}

@@

\begin{figure*}[tbh]
  \centering
  \subfigure[Best case.]{\includegraphics[width=.31\linewidth]{figures/flops_to_bytes_best-crop.pdf}\label{fig:ratio_best}}%
  \subfigure[Worst case.]{\includegraphics[width=.31\linewidth]{figures/flops_to_bytes_worst-crop.pdf}\label{fig:ratio_worst}}%
  \subfigure[Worst case (No cacheline effects)]{\includegraphics[width=.31\linewidth]{figures/flops_to_bytes_no_cache-crop.pdf}\label{fig:ratio_worst_nocache}}

@@

\begin{figure}[t]
  \centering 
  \includegraphics[width=.95\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a bandwidth from memory to computational units of 150~GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
  \vspace{3ex}
\end{figure}

@@

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first array is {\tt col\_id} that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists
of $n_z$ entries that label the column index of each nonzero in
$A$. %***
The other array is  {\tt data} that gives the values of
the nonzeros in the four matrices. The values of the matrices are
interleaved by groups of four so that the value $A^l_{col\_id[k]}$ of
the $k$th nonzero of the $l$th matrix is at index $\floor{\frac{k-1}{4}}*4*n_m + (l-1)*4+(k-1)\%4$.
(In other words, the first four entries are from $A^{1}$ and the 
next four are from $A^2$.) The vectors {\tt x} are
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly 
grouped in sets of 16, which correspond to four derivatives applied to
four functions, at a single point. %\ger{All matrices are stored as one-dimensional vectors and pointers to individual rows and elements are explicitely computed.}
%GE \looseness=-1

@@

  \caption{Bandwidth performance under idealized conditions.  Entries
    with "cpp" (triangles) denote cases where vectorization is
    compiler generated.}
  \label{fig:bandwidth}
  \label{fig:band_rw_MIC}
  \vspace{2ex}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=.9\linewidth]{figures/bandwidth_gather_unpack.pdf}
  \caption{Performance of {\tt gather} and {\tt unpack} operations.}
  \label{fig:band_gather} 
  \vspace{2ex}
\end{figure}

@@

\begin{figure*}
% Figure 9
  \centering
% was width=.48
  \subfigure[Supercompact and Random.]{\includegraphics[width=.45\linewidth]{figures/random_supercompact.pdf}\label{fig:perf_ideal}}
% was width=.48
  \subfigure[Compact and RBF-FD.]{\includegraphics[width=.45\linewidth]{figures/rbf_compact.pdf}\label{fig:perf_realistic}}

@@

  \includegraphics[width=.9\linewidth]{figures/mic_performance_nb_threads.pdf}

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical "one-matrix one-vector" SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.
    We also show peak theoretical performance for v1m1 and v4m1 on the MIC. 
    Not shown is the peak theoretical performance of 210 Gflops for v4m4.  }
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
  \vspace{2ex}
\end{figure}

@@

\section*{Acknowledgment}
Erlebacher was supported under NSF grant
DMS-\#0934331 (FSU), and the use of the MIC at FSU and at LCSE/U. Minnesota (provided by Prof. D. Yuen). NCAR is sponsored by NSF. 
Flyer acknowledges support of NSF grant DMS-0934317.
